Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-03-12 Manuscript/HI_Model_ms.Rmd
updated manuscript files

## 1. Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and angricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural carbon solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-effective and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, the growing number of sub-national jurisdictions that plan to incorporate land-based mitigation strategies into their emissions reduction efforts would benefit from understanding how future land use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands on an annual basis for the period 2010–2100 under a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. We explored four unique scenarios that represent all combinations of two land use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. our goals were to estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and assess the sensitivity of model estimates to the introduction of a CFE on NPP.

@@

### *2.4 CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@Walker_integrating_2020]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@Walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model projections of ecosystem carbon balance to different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

@@

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2020 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019]. In contrast, our NBP estimate for the past decade was ~ 88% higher than a previous statewide LUCAS model estimate covering the same time period [0.341 Tg C y^-1^; @selmants_baseline_2017]. This discrepancy was likely driven by modifications in how we calculated NPP, soil R~h~, and soil aquatic carbon loss compared to previous versions of the LUCAS model, as well our modelʻs finer spatial resolution [@selmants_baseline_2017; @daniel_integrating_2018]. Previous versions of a Hawaiʻi LUCAS model were run at 1-km spatial resolution and simulation cells within each unique combination of moisture zone and state type all had the same mean IBIS-derived NPP value applied to them at the beginning of the simulation period. In contrast, our NPP estimates at 250-m spatial resolution were adjusted on a cell-by-cell basis using Hawaiʻi-specific climate data as described in section 2.3. As a result, our statewide NPP estimates from 2011-2020 were 9.5% lower on average than previous LUCAS model estimates for Hawaiʻi during the same time period [@selmants_baseline_2017], likely because of the greater influence of more arid simulation cells. Soil carbon losses via R~h~, leaching, and overland flow in previous versions of the LUCAS model were calculated as the ratio of the IBIS-derived flux to the size of the originating carbon stock, in this case soil organic carbon to 1-m depth [@daniel_integrating_2018]. Here we calculated soil R~h~ and aquatic carbon losses as the ratio of the mean IBIS-derived flux to the microclimate-adjusted NPP value of each simulation cell, which is a more realistic driver than stock size [@jackson_ecology_2017]. Compared to previous Hawaiʻi LUCAS model estimates [@selmants_baseline_2017], soil R~h~ and aquatic carbon losses from 2011-2019 were reduced by an average of 15% and 21%, respectively, which widened the gap between between carbon gain (NPP) and carbon losses and accounted for the overall increase in NBP estimates for this time period.

### *4.3 Limitations of this study*

>>>>

## 1. Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and agricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural carbon solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-efficient and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, the growing number of sub-national jurisdictions that plan to incorporate land-based mitigation strategies into their emissions reduction efforts would benefit from understanding how future land use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands on an annual basis for the period 2010–2100 under a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. We explored four unique scenarios that represent all combinations of two land use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and RCP 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. our goals were to estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and assess the sensitivity of model estimates to the introduction of a CFE on NPP.

@@

### *2.4 CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@franks_sensitivity_2013]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model projections of ecosystem carbon balance to different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

@@

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2020 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 Tg C in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019]. In contrast, our NBP estimate for the past decade was ~ 88% higher than a previous statewide LUCAS model estimate covering the same time period [0.341 Tg C y^-1^; @selmants_baseline_2017]. This discrepancy was likely driven by modifications in how we calculated NPP, soil R~h~, and soil aquatic carbon loss compared to previous versions of the LUCAS model, as well our modelʻs finer spatial resolution [@selmants_baseline_2017; @daniel_integrating_2018]. Previous versions of a Hawaiʻi LUCAS model were run at 1-km spatial resolution and simulation cells within each unique combination of moisture zone and state type all had the same mean IBIS-derived NPP value applied to them at the beginning of the simulation period. In contrast, our NPP estimates at 250-m spatial resolution were adjusted on a cell-by-cell basis using Hawaiʻi-specific climate data as described in section 2.3. As a result, our statewide NPP estimates from 2011-2020 were 9.5% lower on average than previous LUCAS model estimates for Hawaiʻi during the same time period [@selmants_baseline_2017], likely because of the greater influence of more arid simulation cells. Soil carbon losses via R~h~, leaching, and overland flow in previous versions of the LUCAS model were calculated as the ratio of the IBIS-derived flux to the size of the originating carbon stock, in this case soil organic carbon to 1-m depth [@daniel_integrating_2018]. Here we calculated soil R~h~ and aquatic carbon losses as the ratio of the mean IBIS-derived flux to the microclimate-adjusted NPP value of each simulation cell, which is a more realistic driver than stock size [@jackson_ecology_2017]. Compared to previous Hawaiʻi LUCAS model estimates [@selmants_baseline_2017], soil R~h~ and aquatic carbon losses from 2011-2020 were reduced by an average of 15% and 21%, respectively, which widened the gap between between carbon gain (NPP) and carbon losses and accounted for the overall increase in NBP estimates for this time period.

### *4.3 Limitations of this study* 
There is ample evidence that increasing atmospheric CO~2~ concentrations can stimulate NPP [@norby_forest_2005; @zhu_greening_2016], but the magnitude and persistence of this effect across climatic and nutrient availability gradients remains highly uncertain, especially over long time spans [@franks_sensitivity_2013; @walker_integrating_2020]. Our results demonstrate that long-term projections of ecosystem carbon balance are highly sensitive to uncertainty in the strength of a CFE. With no CFE, Hawaiian ecosystems became a net source of CO~2~ to the atmosphere beginning in the latter half of the 21st century under the high land use and high radiative forcing scenario. However, a CFE equivalent to a 5% increase in NPP for every 100 ppm increase in atmospheric CO~2~ applied to the same scenario resulted in Hawaiian ecosystems collectively remaining a net carbon sink throughout the entire simulation period. A CFE of 15% applied to the high land use and high radiative forcing scenario resulted in a nearly 5-fold increase in mean annual NBP averaged across all years and Monte Carlo realizations.
